Neighborhoods and Felony Disenfranchisement: The Case of New York City

### 2019-05-18 markdown/kmorris_thesis.Rmd
more writing

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "April 22, 2019"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
link-citations: yes
---

@@

These findings are not suprising given the large body of research establishing the social nature of voting. Outside the felony disenfranchisement conversation, an enormous body of research has demonstrated that individuals are far more likely to cast a ballot if their peers, family members or neighbors are doing so. [Build this out]

@@

# Framing for Turnout Effects {-}

@@

Of course, despite dire over-incarceration in some neighborhoods, the number of incarcerated individuals is relatively low compared to the number of voters. In New York, for instance, `r comma(readRDS("./temp/number_incarcerated.rds"))` individuals were imprisoned in New York State in early 2019, compared with `r round(readRDS("./temp/active_voters_0319.rds")/1000000, digits = 1)` million actively registered voters. Despite the lower share of residents who are disenfranchised through imprisonment, there is reason to believe the policy impacts more individuals than just those imprisoned. As discussed above, previous research has demonstrated that felony disenfranchisement reduces turnout even among Black voters whose rights are not stripped. This research has found, in particular, that eligible Black voters are less likely to cast a ballot in states where felony disenfranchisement policies are harsher. Understanding the distribution of this effect is important for understanding its political implications. Are voters who live in the same neighborhood as the disenfranchised less likely to cast a ballot because their neighbors have been removed from the community? Or does felony disenfranchisement operate cast its chilling effect on turnout more broadly, crossing even into neighborhoods where voters have not been removed from the community?

This paper begins by examining whether felony disenfranchisement is in fact associated with lower turnout rates in a given neighborhood.^[As discussed below, "neighborhood" is difficult to define. I explore turnout differentials at both the census block group and census tract level.] As discussed above, this depressive effect on turnout is expected to operate through two distinct mechanisms: the actual incarceration and disenfranchisment of convicted individuals, and the social effects on eligible voters identified in previous research. These turnout effects are estimated in various ways: by comparing neighborhoods where voters with a history of participating have been incarcerated with similar neighborhoods where voters have not been incarcerated; by using regression analysis to investigate whether neighborhood arrests are correlated with turnout rates; and by using regression analysis to investigate the relationship between prison admissions and turnout at the zip-code level.

# What This Paper Examines {-}
This paper begins by examining the effect of felony disenfranchisement in the New York City Mayoral election of 2017. This paper is primarily interested in whether felony disenfranchisement has impacts on local political distribution. If, for instance, voters in particular marked neighborhoods turnout at lower rates because of disenfranchisement policies, this could have ramifications for city-wide policy. @Martin2013, for instance, documents the relationship between voter turnout and the responsiveness of Congressional representatives. "[H]igher citizen participation is rewarded," they argue, "with enhanced policy responsiveness." If the neighborhoods most impacted by felony disenfranchisement turn out at lower rates, they may find that their elected representatives are less likely to support their needs. @Griffin2005 find similar effects in the United States Senate, reporting that "voter preferences predict the aggregate roll-call behavior of Senators while nonvoter preferences do not."

Although research on the impact of local turnout on city-wide policy is scarce, @Anzia2019 examines the impact of senior turnout on "senior-friendly" policy at the city level. Anzia does not find that senior turnout in general increases the likelihood of senior-friendly policies, but that elected officials *are* responsive to senior turnout when seniors "are a more cohesive, meaningful group." American cities are highly segregated by race and by class, but less so by age. As such, Anzia's study does not speak directly to the impact of *neighborhood* turnout rates on city policy. An extension of her paper, however, implies that what she finds might be probitive to the effect of neighborhood turnout rates. Insofar as neighborhoods vote cohesively on issues such as school and park funding, their turnout rates may influence city policy.

After examining the effect of felony disenfranchisement on voter turnout at the neighborhood level, I consider the efficacy of a program intended to undo the impact of felony disenfranchisement. In April of 2018, Governor Andrew Cuomo signed Executive Order 181 which restored voting rights to New Yorkers on parole.^[Prior to Executive Order 181, residents in prison and on parole were barred from voting, while individuals sentenced to probation did not have their voting rights revoked. Today, only those in prison are disenfranchised] Parole officers were required to provide registration forms the parolees under their supervision and inform them of their voting rights. In theory, such a policy could increase turnout and registration for these individuals. To test whether the policy was successful, I compare the turnout rates in the 2018 general election of three groups who finished parole in 2018: those who finished parole immediately before the policy change, those who finished parole immediately after the policy change, and those who finished parole immediately after the election. These three groups allow me to explore whether the re-extension of voting rights while on parole has an impact on turnout (by comparing the first two groups) and whether being under formal supervision while being allowed to vote results in higher turnout than being allowed to vote but being under supervision (by comparing the third group to the first two).

@@

The state does not make a unified database of parolees whose voting rights have been restored available to the public. However, the Department of Corrections and Community Supervision website's Parolee Lookup includes a flag indicating whether someone's voting rights haave been restored. By using identification numbers from the data obtained from the state to lookup parolees on the website, I was able to construct a list of the individuals who were on parole and had their rights restored.

>>>>

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "April 22, 2019"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
link-citations: yes
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "April 22, 2019" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->

@@

@@

# What This Paper Examines {-}

This paper begins by examining the effect of felony disenfranchisement in the New York City Mayoral election of 2017. I am primarily interested in whether felony disenfranchisement has impacts on local political distribution. As discussed above, policing and incarceration patterns have historically targeted communities of color, wreaking havok on the social fabric of these neighborhoods. It is possible, however, that these effects are even broader than the direct focus on policing has acknowledged. We know that felony disenfranchisement systematically removes voters from certain neighborhoods, but it is not clear whether enough voters are removed relative to the electorate to meaningfully distort neighborhood representation. However, to the extend that felony disenfranchisement has a depressive effect on *eligible* voters, it is possible that disenfranchisement policies are powerful enough to materially reduce the representatition of certain parts of the city. Although turnout is not a perfect proxy for political representation, there is some evidence that policians are more responsive to groups that turnout at higher rates. @Martin2013, for instance, documents the relationship between voter turnout and the responsiveness of Congressional representatives. "[H]igher citizen participation is rewarded," they argue, "with enhanced policy responsiveness." If the neighborhoods most impacted by felony disenfranchisement turn out at lower rates, they may find that their elected representatives are less likely to support their needs. @Griffin2005 find similar effects in the United States Senate, reporting that "voter preferences predict the aggregate roll-call behavior of Senators while nonvoter preferences do not." If felony disenfranchisement has a depressive effect on turnout, it is likely that it is responsible for politicians being less responsive to the particular preferences of voters in neighborhoods where voters are removed --- some of the most disadvantaged neighborhoods to begin with.

After examining the effect of felony disenfranchisement on voter turnout at the neighborhood level, I consider the efficacy of a program intended to undo the impact of felony disenfranchisement. In April of 2018, Governor Andrew Cuomo signed Executive Order 181 which restored voting rights to New Yorkers on parole.^[Prior to Executive Order 181, residents convicted of felonies in prison and on parole were barred from voting, while individuals sentenced to probation did not have their voting rights revoked. Today, only those in prison for felony offenses are disenfranchised.] Parole officers were required to provide registration forms the parolees under their supervision and inform them of their voting rights. In theory, such a policy could increase turnout and registration for parolees. To test whether the policy was successful, I compare the turnout rates in the 2018 general election of three groups who finished parole in 2018: those who finished parole immediately before the policy change, those who finished parole immediately after the policy change, and those who finished parole immediately after the election. These three groups allow me to explore whether the re-extension of voting rights while on parole has an impact on turnout (by comparing the first two groups) and whether being under formal supervision while being allowed to vote results in higher turnout than being allowed to vote but being under supervision (by comparing the third group to the first two).

The first part of this project seeks to refine how we think about felony disenfranchisement in the United States. Rather thank think about it purely through the prism of those directly impacted (residents who are actively disenfranchised) or through a prism of race (by examining the impact of disenfranchisement on African-American turnout at the state level), I seek to introduce the importance of physical space into the conversation. This will allow us to both better understand the mechanisms through which the depressive effects operate and also to understand more deeply where these effects are concentrated. Of course, understanding the impact of felony disenfranchisement is not enough: we must also be testing whether our efforts to undo its legacy are effective. The second part of this project seeks to do just that.

## Framing for Turnout Effects {-}

@@

Despite dire over-incarceration in some neighborhoods, the number of incarcerated individuals is relatively low compared to the number of voters. In New York, for instance, `r comma(readRDS("./temp/number_incarcerated.rds"))` individuals were imprisoned in New York State in early 2019, compared with `r round(readRDS("./temp/active_voters_0319.rds")/1000000, digits = 1)` million actively registered voters. Despite the low share of residents who are directly disenfranchised, there is reason to believe the policy impacts more individuals than just those imprisoned. As discussed above, previous research has demonstrated that felony disenfranchisement reduces turnout even among Black voters whose rights are not stripped. This research has found, in particular, that eligible Black voters are less likely to cast a ballot in states where felony disenfranchisement policies are harsher, an effect often referred to as *de facto* disenfranchisement. The existence of such *de facto* disenfranchisement is bolstered by a large body of empirical work on the social nature of voting. Many scholars have established that being in a community where voting is common makes individuals more likely to vote. Children whose parents vote, for instance, are more likely to vote as adults [XX]; XX add more examples here. Many of these relationships are spatially bounded: family members are likely to live in the same home, or on the same block. Literature from urban sociology has established that social networks are largely spatially bounded. To the extent that felony disenfranchisement policies have depressive effects on turnout in the social and filial networks of the imprisoned, these policies are likely to be closely concentrated in the neighborhoods where the disenfranchised live.

Although research on the impact of local turnout on city-wide policy is scarce, @Anzia2019 examines the impact of senior turnout on "senior-friendly" policy at the city level. Anzia does not find that senior turnout in general increases the likelihood of senior-friendly policies, but that elected officials *are* responsive to senior turnout when seniors "are a more cohesive, meaningful group." American cities are highly segregated by race and by class, but less so by age. As such, Anzia's study does not speak directly to the impact of *neighborhood* turnout rates on city policy. An extension of her paper, however, implies that what she finds might be probitive to the effect of neighborhood turnout rates. Insofar as neighborhoods vote as a bloc, their turnout rates may influence city policy.

## Previous Attempts to Re-Incorporate Formerly Disenfranchised Residents

@@

The state does not make a unified database of parolees whose voting rights have been restored available to the public. However, the Department of Corrections and Community Supervision website's Parolee Lookup includes a flag indicating whether someone's voting rights haave been restored. By using identification numbers from the data obtained from the state to lookup parolees on the website, I was able to construct a list of the individuals who were on parole and had their rights restored.^[This list was compiled by using a webscraper written in the Python language.]
